Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-08 WSSSPE1_report/WSSSPE1_summary_report.tex
minor changes in a few places

Additional volunteers:
\begin{itemize}
\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
\item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
\item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
\item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
\item Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
\item Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
\item * Frank Löffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
\item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
\item James Howison $\langle$\url{james@howison.name}$\rangle$
\item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
\item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
\item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
\item Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
\end{itemize}

@@

Because progress in scientific research is dependent on the quality and
accessibility of software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.\note{Ketan: What are the levels in scientific process where software
is required. It might be interesting to outline these stages and types of
software required in each stage.} In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and discussion have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
journal publication.  Additional papers based on extended versions of workshop
submissions are expected.  The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC conferences) and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress.  A minisymposium on ``Reliable Computational
Science'' that aims to further explore some of the key issues raised in WSSSPE
is co-organized by a participant in the 2014 SIAM Annual Meeting.

@@

Bourne's keynote presentation is entitled ``A Recipe for Sustainable
Software''.  He emphasized that sustainability for software ``does not
just mean more money from Government.''  He noted that in places,
development in science has improved thanks to open source and Github,
but for the most part remains arcane. He argued that there is much we
can learn from the App Store model. He mentioned
BioJava~\cite{biojava-web} and OpenScienceDataCloud~\cite{osdc-web} as
distribution channels.  On a related note, Bourne observed a common
evolutionary pathway for computational biology projects, from data
archive to analytics platform to educational use, and suggested that
use of scientific software for outreach might be the final step.

Bourne shared with the audience a few real challenges he
encountered. First, he has looked into reproducibility in
computational biology, and has concluded that ``I have proved I cannot
reproduce research from my own lab''~\cite{Veretnik}.

Another problem he experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of private sector sustainability sources and public-private
partnerships. He noted that making a successful business from
scientific software alone is very rare: founders overvalue while
customers undervalue. He noted that to last, an open source project
needs a minimal funding requirement even with a vibrant community --
goodwill only goes so far if you are being paid to do something else.
He talked about grant schemes of relevance in the U.S., particularly
with regard to technology transfer~\cite{sbir-web, fased-web}.

@@

One of the paper by Maheshwari et al~\cite{Maheshwari_WSSSPE} that was
discussed was about the role of technological catalysts in overall e-research
process.  The paper focuses on ``technology catalysts" and their role in modern
scientific research process. Technology catalyst could be defined as a role
played by an individual with a knowhow of technological advancements tasked
with user engagement with a goal of enacting scientific or engineering
applications using suitable tools and techniques to take advantage of
technological capabilities and benefitting applications.

@@

Nancy Wilkins-Diehr summarized 6 papers in "Communities Part 2". The Christopherson paper outlines the degree to which research
relies on high quality software. There are often barriers and a lack of suitable incentives for researchers to embrace software engineering principals.
The Water Science Software Institute is working to lower some of these barriers through an Open Community Engagement Process.
This is a 4-step iterative development process that incorporates Agile development principals.

@@

The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hack-athon to develop working code.
Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of
code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.

>>>>

Additional volunteers:
\begin{itemize}
\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
\item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
\item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
\item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
\item Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
\item * Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
\item * Frank L\"{o}ffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
\item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
\item James Howison $\langle$\url{james@howison.name}$\rangle$
\item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
\item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
\item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
\item Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
\end{itemize}

@@

Because progress in scientific research is dependent on the quality and
accessibility of software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and discussion have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
journal publication.  Additional papers based on extended versions of workshop
submissions are expected.  The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC conferences) and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress.  A minisymposium on ``Reliable Computational
Science'' that aims to further explore some of the key issues raised in WSSSPE
is co-organized by a participant in the 2014 SIAM Annual Meeting.

@@

Bourne's keynote presentation was entitled ``A Recipe for Sustainable
Software.''  He emphasized that sustainability for software ``does not
just mean more money from Government.''  He noted that in places,
development in science has improved thanks to open source and Github,
but for the most part remains arcane. He argued that there is much we
can learn from the App Store model. He mentioned
BioJava~\cite{biojava-web} and OpenScienceDataCloud~\cite{osdc-web} as
distribution channels.  On a related note, Bourne observed a common
evolutionary pathway for computational biology projects, from data
archive to analytics platform to educational use, and suggested that
use of scientific software for outreach might be the final step.

Bourne shared with the audience a few real challenges he
encountered. First, he has looked into reproducibility in
computational biology, and has concluded that ``I have proved I cannot
reproduce research from my own lab.''~\cite{Veretnik}

Another problem he experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of private sector sustainability sources and public-private
partnerships. He noted that making a successful business from
scientific software alone is very rare: founders overvalue while
customers undervalue. He noted that to last, an open source project
needs a minimal funding requirement even with a vibrant community --
goodwill only goes so far if you are being paid to do something else.
He talked about grant schemes of relevance in the U.S., particularly
with regard to technology transfer.~\cite{sbir-web, fased-web}

@@

One paper by Maheshwari et al.~\cite{Maheshwari_WSSSPE} that was
discussed was about the role of technological catalysts in overall e-research
process.  The paper focuses on ``technology catalysts" and their role in modern
scientific research process. Technology catalyst could be defined as a role
played by an individual with a knowhow of technological advancements tasked
with user engagement with a goal of enacting scientific or engineering
applications using suitable tools and techniques to take advantage of
technological capabilities and benefitting applications.

@@

Nancy Wilkins-Diehr summarized 6 papers in ``Communities Part 2.'' The Christopherson paper outlines the degree to which research
relies on high quality software. There are often barriers and a lack of suitable incentives for researchers to embrace software engineering principals.
The Water Science Software Institute is working to lower some of these barriers through an Open Community Engagement Process.
This is a 4-step iterative development process that incorporates Agile development principals.

@@

The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hack-athon \todo{we spell hackathon three different ways in two paragraphs - Pick one and be consistent.} to develop working code.
Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of
code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.
